Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2014-12-21 papers/wdmerger_I/paper.tex
Added a subsection on our uniform sphere and cube convergence test for the Poisson solver.

A related issue that required a code improvement was that in cases of large density gradients such as the edge of a star, it is possible to generate negative densities in zones even with the more strongly limited PPM. This can occur if a region of large density is moving away from an ambient zone at relatively large speeds; then the net density flux in the ambient zones can be large enough to unphysically drag the density below zero. In practice, this will occur at the trailing edge of a star that is moving across a grid. In such a situation, there are two main approaches one could take: either explicitly introduce a positivity-guaranteeing diffusive flux, or reset the characteristics of the affected zone. We choose the latter approach. Even though it is non-conservative, it preserves a characteristic we value, which is to keep the edge of the stars relatively sharp, as they physically should be. Since the mass of the affected zones is typically already fairly low, this should not seriously affect the energy conservation properties of our simulation. Similarly, it is also possible in this physical situation for very large velocities to be generated in the material behind the star. The momentum transferred to the ambient zone is close to the momentum of the stellar material, but this is of much higher density and so the velocity of the ambient material must become very large to compensate. While this may be physically viable in certain circumstances, it poses a problem for our code, which uses subcycling in time (see below). The timestep is fixed at the beginning of the set of subcycled timesteps, and if in the middle of the cycle a velocity is generated that is so large that it violates the CFL criterion for this timestep, we cannot guarantee the stability of the resulting solution since we cannot alter the timestep. (Nor would we want to, because in some cases because the velocity becomes spuriously large and would drag down the global timestep.) Therefore we also insert a check at the end of every hydrodynamics timestep that resets the velocity and/or sound-speed (indirectly, using the temperature) of any zone that violates the CFL criterion given the timestep we chose at the beginning of the cycle. In practice, we find that this is mainly necessary only for simulations in the inertial reference frame of the white dwarf merger, which we do not normally do for the reasons stated in Section \ref{sec:Kepler}.

@@

For the evolution of binary systems, it is most natural to evolve the two stars in a frame that is co-rotating at the same period as the orbital period. CASTRO has the ability to evolve systems in a rotating reference frame, as discussed in the original code paper. Source terms corresponding to the Coriolis and centrifugal force terms are added to the momentum and energy equations. In this frame, the stars essentially remain stationary in their original positions due to the centrifugal force supporting against the gravitational attraction, and will remain this way as long as significant mass transfer does not occur. \cite{swesty00} demonstrated (in the context of neutron star mergers) that conservation of angular momentum is much easier to obtain in the rotating reference frame than in an inertial frame in which stars advect large amounts of material around the domain. We wish to emphasize that although it is commonly stated in the literature that grid-based codes poorly conserve angular momentum, this is not generally true. When the resolution is sufficiently high and a rotating reference frame is employed, excellent conservation properties can result, as demonstrated in Section \ref{sec:Kepler}. We note that as the stars begin to coalesce, the rotating reference frame will no longer provide a good approximation to the spatial motion of the stars and then they will begin to significantly move around the domain. This is not necessarily problematic because the most important feature of the rotating frame is that it helps ensure that the initial coalescence is not the result of spurious numerical loss of angular momentum. When significant mass transfer sets in and evolution proceeds on a dynamical timescale, the conservation properties may be slightly worse but angular momentum conservation is also less important.

@@

\subsubsection{Boundary Conditions}

Analytic solutions to the Poisson equation customarily assume that the potential vanishes at large distances from the region of non-zero density. However, on a finite computational domain it is usually not possible to have the edges of the domain be far enough away that the potential can be taken to be zero there. Solving the Poisson equation therefore requires knowledge of the values of the potential on the edges of the computational domain. In principle, they can be computed by doing a direct sum over the mass distribution inside the domain, where the mass in each zone is treated as a point source:
\begin{equation}
  \Phi_{\text{lmn}} = G\, \sum_{\text{i, j, k}} \frac{G \rho_{\text{ijk}}}{|\mathbf{x}_{\text{lmn}} - \mathbf{x}_{\text{ijk}}|}\, \Delta V_{\text{ijk}}.\label{direct_sum}
\end{equation}
Here (i, j, k) are the indices of cells inside the domain, and (l, m, n) are the indices of boundary locations. $\Delta V$ is the volume of the zone. We have implemented this as an option\footnote{It is controlled with the \texttt{gravity.direct\_sum\_bcs} input parameter.} in CASTRO. If there are $N$ zones per spatial dimension, then there are $6 N^2$ boundary zones, and each boundary zone requires a sum over $N^3$ zones, so the direct computation of the boundary conditions scales as $N^5$.  This method is expensive enough that it is not used for hydrodynamics simulations (though it is useful for comparison to approximate solutions).

In a typical simulation we place the boundaries of the domain far enough away from the region containing most of the mass that some method of approximation to this direct summation is justified. Many approaches exist in the literature. The original release of CASTRO featured the crudest possible approximation: a monopole prescription, where the boundary values were computed by summing up all the mass on the domain and treating it as a point source at the domain center. This is exactly correct only for a spherically symmetric mass distribution, and indeed this is often a sufficient approximation for single-star Type Ia supernova simulations that employ self-gravity (ref). However, for a problem like that of a binary star system with significant departures from spherical symmetry, this assumption fails to produce accurate boundary values. This results in a significant drift of the center of the mass of the system over time.

The most natural extension of the monopole prescription is to include higher-order multipole moments. If the entire mass distribution is enclosed, then the potential can be expanded in a series of spherical harmonics $Y_{lm}$:
\begin{equation}
  \Phi(\mathbf{x}) = \sum_{l=0}^{\infty}\sum_{m=-l}^{l} \frac{4\pi}{2l + 1} q_{lm} \frac{Y_{lm}(\theta,\phi)}{r^{l+1}}, \label{spherical_harmonic_expansion}
\end{equation}
where $q_{lm}$ are the so-called multipole moments. The origin of the coordinate system is taken to be the center of the computational domain, and $r$ is the distance to the origin. The multipole moments can be calculated by expanding the Green's function for the Poisson equation as a series of spherical harmonics.
\begin{comment}
, which yields
\begin{equation}
  q_{lm} = \int Y^*_{lm}(\theta^\prime, \phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3x^\prime. \label{multipole_moments_original}
\end{equation}
\end{comment}
After some algebraic simplification of Equation \ref{spherical_harmonic_expansion} using the addition theorem for spherical harmonics,
\begin{comment}
\begin{align}
  &\frac{4\pi}{2l+1} \sum_{m=-l}^{l} Y^*_{lm}(\theta^\prime,\phi^\prime)\, Y_{lm}(\theta, \phi) = P_l(\text{cos}\, \theta) P_l(\text{cos}\, \theta^\prime) \notag \\
  &\ \ + 2 \sum_{m=1}^{l} \frac{(l-m)!}{(l+m)!} P_{l}^{m}(\text{cos}\, \theta)\, P_{l}^{m}(\text{cos}\, \theta^\prime)\, \left[\text{cos}(m\phi)\, \text{cos}(m\phi^\prime) + \text{sin}(m\phi)\, \text{sin}(m\phi^\prime)\right].
\end{align}
\end{comment}
the potential outside of the mass distribution can be written as:
\begin{equation}
  \Phi(\mathbf{x}) = \sum_{l=0}^{\infty} \left[Q_l^{(0)} \frac{P_l(\text{cos}\, \theta)}{r^{l+1}} + \sum_{m = 1}^{l}\left[ Q_{lm}^{(C)}\, \text{cos}(m\phi) + Q_{lm}^{(S)}\, \text{sin}(m\phi)\right] \frac{P_{l}^{m}(\text{cos}\, \theta)}{r^{l+1}} \right].\label{multipole_potential}
\end{equation}
$P_l(x)$ are the Legendre polynomials and $P_{l}^{m}(x)$ are the associated Legendre polynomials. $Q_l^{(0)}$ and $Q_{lm}^{(C,S)}$ are variants of the multipole moments that involve integrals of $P_l$ and $P_l^m$, respectively, over the computational domain.
\begin{comment}
\begin{align}
  Q_l^{(0)}   &= \int P_l(\text{cos}\, \theta^\prime)\, {r^{\prime}}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(C)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{cos}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(S)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{sin}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime.
\end{align}
\end{comment}
This approach becomes computationally feasible when we cut off the outer summation in Equation \ref{multipole_potential} at some finite value of $l_{\text{max}}$. If it is of sufficiently high order, we will accurately capture the distribution of mass on the grid. In practice we first evaluate the discretized analog of the modified multipole moments for $0 \leq l \leq l_{\text{max}}$ and $1 \leq m \leq l$, an operation that scales as $N^3$. We then directly compute the value of the potential on all of the $6N^2$ boundary zones. Since the multipole moments only need to be calculated once, the full operation scales only as $N^3$. The amount of time required to calculate the boundary conditions will be directly related to the chosen value of $l_{\text{max}}$, so there is a trade-off between computational expense and accuracy of the result. We find that for the early stages of the evolution of the binary system, $l_\text{max} \approx 6$ is sufficient.

@@

A simple test to verify the Poisson solver implemented by CASTRO is
the case of gravitational free fall. In this setup, two stars, each
individually in an equilibrium state, are placed on the computational
grid, separated by an initial distance $r_0$ along the $x$ axis with
zero initial velocity. We choose stars of masses $0.6$ and $0.8\,
M_\odot$, with the lower mass star on the left side (e.g. $x < 0$) of
the domain and the higher mass star on the right, such that their
center of mass coincides with the center of the
domain. Gravitationally, the stars may be treated as point masses
until the point of contact, so the equation of motion governing the
distance $r$ between their centers of mass is that of simple free
fall:
\begin{equation}
  \ddot{r}(t) = - \frac{GM}{r},
\end{equation}
where $G$ is the gravitational constant and $M$ is the total mass of
the system. This differential equation has a closed-form solution for
the evolution time as a function of separation:
\begin{equation}
  t(r) = \sqrt{\frac{r_0^3}{2GM}} \left[ \text{arccos}\left(\sqrt{\frac{r}{r_0}}\,\right) + \sqrt{\frac{r}{r_0} \left(1 - \frac{r}{r_0}\right)}\ \right]. \label{analyticalFreeFall}
\end{equation}\MarginPar{This result is derived in the freefall directory.}
This result can be derived by recognizing that
\[
  t(r) = \int_{r_0}^r \frac{dr}{v(r)}
\]
and inserting the velocity as a function of radial separation,
\[
  v(r) = \sqrt{\frac{2GM}{r_0}\left(\frac{r_0}{r} - 1\right)}.
\]
We determine the initial separation to be consistent with the
simulation performed in Section \ref{sec:Kepler}; that is, we select
an initial orbital period and use Kepler's third law to calculate the
radius of a circular orbit with that period. We select an initial
orbital period of $T = 100$ s for this simulation, so that the separation is
\[
  r_0 = 3.61 \times 10^{9}\ \text{cm}.
\]
We chose a relatively low resolution simulation to demonstrate the
capabilities of CASTRO even while using modest resources. The
computational grid is covered by a coarse grid of $48^3$ zones, with
two levels of refinement above the coarse grid. Each refined grid
carries an increase in resolution by a factor of 4 relative to the
coarser grid below it. CASTRO initially assigns $93\%$ of the domain
to be covered by the intermediate resolution grids, and $0.05\%$ of
the domain to be covered by the finest resolution grids.

@@

\subsection{Keplerian Orbit}\label{sec:Kepler}

>>>>

A related issue that required a code improvement was that in cases of large density gradients such as the edge of a star, it is possible to generate negative densities in zones even with the more strongly limited PPM. This can occur if a region of large density is moving away from an ambient zone at relatively large speeds; then the net density flux in the ambient zones can be large enough to unphysically drag the density below zero. In practice, this will occur at the trailing edge of a star that is moving across a grid. In such a situation, there are two main approaches one could take: either explicitly introduce a positivity-guaranteeing diffusive flux, or reset the characteristics of the affected zone. We choose the latter approach. Even though it is non-conservative, it preserves a characteristic we value, which is to keep the edge of the stars relatively sharp, as they physically should be. Since the mass of the affected zones is typically already fairly low, this should not seriously affect the energy conservation properties of our simulation. Similarly, it is also possible in this physical situation for very large velocities to be generated in the material behind the star. The momentum transferred to the ambient zone is close to the momentum of the stellar material, but this is of much higher density and so the velocity of the ambient material must become very large to compensate. While this may be physically viable in certain circumstances, it poses a problem for our code, which uses subcycling in time (see below). The timestep is fixed at the beginning of the set of subcycled timesteps, and if in the middle of the cycle a velocity is generated that is so large that it violates the CFL criterion for this timestep, we cannot guarantee the stability of the resulting solution since we cannot alter the timestep. (Nor would we want to, because in some cases because the velocity becomes spuriously large and would drag down the global timestep.) Therefore we also insert a check at the end of every hydrodynamics timestep that resets the velocity and/or sound-speed (indirectly, using the temperature) of any zone that violates the CFL criterion given the timestep we chose at the beginning of the cycle. In practice, we find that this is mainly necessary only for simulations in the inertial reference frame of the white dwarf merger, which we do not normally do for the reasons stated in Section \ref{sec:kepler}.

@@

For the evolution of binary systems, it is most natural to evolve the two stars in a frame that is co-rotating at the same period as the orbital period. CASTRO has the ability to evolve systems in a rotating reference frame, as discussed in the original code paper. Source terms corresponding to the Coriolis and centrifugal force terms are added to the momentum and energy equations. In this frame, the stars essentially remain stationary in their original positions due to the centrifugal force supporting against the gravitational attraction, and will remain this way as long as significant mass transfer does not occur. \cite{swesty00} demonstrated (in the context of neutron star mergers) that conservation of angular momentum is much easier to obtain in the rotating reference frame than in an inertial frame in which stars advect large amounts of material around the domain. We wish to emphasize that although it is commonly stated in the literature that grid-based codes poorly conserve angular momentum, this is not generally true. When the resolution is sufficiently high and a rotating reference frame is employed, excellent conservation properties can result, as demonstrated in Section \ref{sec:kepler}. We note that as the stars begin to coalesce, the rotating reference frame will no longer provide a good approximation to the spatial motion of the stars and then they will begin to significantly move around the domain. This is not necessarily problematic because the most important feature of the rotating frame is that it helps ensure that the initial coalescence is not the result of spurious numerical loss of angular momentum. When significant mass transfer sets in and evolution proceeds on a dynamical timescale, the conservation properties may be slightly worse but angular momentum conservation is also less important.

@@

\subsubsection{Coupling to Hydrodynamics}\label{sec:gravity_hydro_coupling}

\subsubsection{Boundary Conditions}\label{sec:gravity_boundary_conditions}

Analytic solutions to the Poisson equation customarily assume that the potential vanishes at large distances from the region of non-zero density. However, on a finite computational domain it is usually not possible to have the edges of the domain be far enough away that the potential can be taken to be zero there. Solving the Poisson equation therefore requires knowledge of the values of the potential on the edges of the computational domain. In principle, they can be computed by doing a direct sum over the mass distribution inside the domain, where the mass in each zone is treated as a point source:
\begin{equation}
  \Phi_{\text{lmn}} = G\, \sum_{\text{i, j, k}} \frac{G \rho_{\text{ijk}}}{|\mathbf{x}_{\text{lmn}} - \mathbf{x}_{\text{ijk}}|}\, \Delta V_{\text{ijk}}.\label{direct_sum}
\end{equation}
Here (i, j, k) are the indices of cells inside the domain, and (l, m, n) are the indices of boundary locations. $\Delta V$ is the volume of the zone. We have implemented this as an option \footnote{It is controlled with the \texttt{gravity.direct\_sum\_bcs} input parameter.} in CASTRO. If there are $N$ zones per spatial dimension, then there are $6 N^2$ boundary zones, and each boundary zone requires a sum over $N^3$ zones, so the direct computation of the boundary conditions scales as $N^5$.  This method is expensive enough that it is not used for hydrodynamics simulations (though it is useful for comparison to approximate solutions).

In a typical simulation we place the boundaries of the domain far enough away from the region containing most of the mass that some method of approximation to this direct summation is justified. Many approaches exist in the literature. The original release of CASTRO featured the crudest possible approximation: a monopole prescription, where the boundary values were computed by summing up all the mass on the domain and treating it as a point source at the domain center. This is exactly correct only for a spherically symmetric mass distribution, and therefore is best suited for problems like single-star Type Ia supernova simulations that employ self-gravity. However, for a problem like that of a binary star system with significant departures from spherical symmetry, this assumption fails to produce accurate boundary values. This results in a significant drift of the center of the mass of the system over time, which will be discussed in Section \ref{sec:kepler}.

The most natural extension of the monopole prescription is to include higher-order multipole moments. If the entire mass distribution is enclosed, then the potential can be expanded in a series of spherical harmonics $Y_{lm}$:
\begin{equation}
  \Phi(\mathbf{x}) = \sum_{l=0}^{\infty}\sum_{m=-l}^{l} \frac{4\pi}{2l + 1} q_{lm} \frac{Y_{lm}(\theta,\phi)}{r^{l+1}}, \label{spherical_harmonic_expansion}
\end{equation}
where $q_{lm}$ are the so-called multipole moments. The origin of the coordinate system is taken to be the center of the computational domain, and $r$ is the distance to the origin. The multipole moments can be calculated by expanding the Green's function for the Poisson equation as a series of spherical harmonics.
\begin{comment}
, which yields
\begin{equation}
  q_{lm} = \int Y^*_{lm}(\theta^\prime, \phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3x^\prime. \label{multipole_moments_original}
\end{equation}
\end{comment}
After some algebraic simplification of Equation \ref{spherical_harmonic_expansion} using the addition theorem for spherical harmonics,
\begin{comment}
\begin{align}
  &\frac{4\pi}{2l+1} \sum_{m=-l}^{l} Y^*_{lm}(\theta^\prime,\phi^\prime)\, Y_{lm}(\theta, \phi) = P_l(\text{cos}\, \theta) P_l(\text{cos}\, \theta^\prime) \notag \\
  &\ \ + 2 \sum_{m=1}^{l} \frac{(l-m)!}{(l+m)!} P_{l}^{m}(\text{cos}\, \theta)\, P_{l}^{m}(\text{cos}\, \theta^\prime)\, \left[\text{cos}(m\phi)\, \text{cos}(m\phi^\prime) + \text{sin}(m\phi)\, \text{sin}(m\phi^\prime)\right].
\end{align}
\end{comment}
the potential outside of the mass distribution can be written as:
\begin{equation}
  \Phi(\mathbf{x}) = \sum_{l=0}^{\infty} \left[Q_l^{(0)} \frac{P_l(\text{cos}\, \theta)}{r^{l+1}} + \sum_{m = 1}^{l}\left[ Q_{lm}^{(C)}\, \text{cos}(m\phi) + Q_{lm}^{(S)}\, \text{sin}(m\phi)\right] \frac{P_{l}^{m}(\text{cos}\, \theta)}{r^{l+1}} \right].\label{multipole_potential}
\end{equation}
$P_l(x)$ are the Legendre polynomials and $P_{l}^{m}(x)$ are the associated Legendre polynomials. $Q_l^{(0)}$ and $Q_{lm}^{(C,S)}$ are variants of the multipole moments that involve integrals of $P_l$ and $P_l^m$, respectively, over the computational domain.
\begin{comment}
\begin{align}
  Q_l^{(0)}   &= \int P_l(\text{cos}\, \theta^\prime)\, {r^{\prime}}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(C)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{cos}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(S)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{sin}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime.
\end{align}
\end{comment}
This approach becomes computationally feasible when we cut off the outer summation in Equation \ref{multipole_potential} at some finite value of $l_{\text{max}}$. If it is of sufficiently high order, we will accurately capture the distribution of mass on the grid. In practice we first evaluate the discretized analog of the modified multipole moments for $0 \leq l \leq l_{\text{max}}$ and $1 \leq m \leq l$, an operation that scales as $N^3$. We then directly compute the value of the potential on all of the $6N^2$ boundary zones. Since the multipole moments only need to be calculated once per Poisson solve, the full operation scales only as $N^3$. The amount of time required to calculate the boundary conditions will be directly related to the chosen value of $l_{\text{max}}$, so there is a trade-off between computational expense and accuracy of the result.

% Insert multipole vs. exact calculation here

\subsubsection{Convergence Testing}\label{sec:gravity_convergence_testing}

Since the results of a merger simulation depend strongly on gravity, it is important to check whether proper numerical convergence is achieved for the Poisson solver. To do so, we created a simple test that inserts a sphere of radius $R$ and uniform mass density $\rho$ onto our grid, and used CASTRO to calculate the gravitational potential $\phi$ of this setup. We ensure that $R$ is an integer multiple of the grid spacing, and the center of the sphere is at the origin. The problem domain for our simulations is $[-1.6, 1.6]^3$, and we take $R = 0.5$ and $\rho = 10^3$. The zones with $r > R$ are filled with an ambient material of very low density. We run this problem at multiple resolutions corresponding to jumps by a factor of two. For comparison, at each grid point we evaluate the analytical potential of a uniform sphere, which can be easily determined using Gauss' law:
\begin{equation}
  \phi_{\text{sphere}}(r) = \frac{GM}{r} \times \begin{cases} (3R^2 - r^2)/(2 r^2) & r \leq R \\ 1 & r > R \end{cases},\label{eq:sphere-analytical}
\end{equation}
where $M = 4\pi R^3 / 3$ is the mass of the sphere. Defining the $L^2$ norm of a field $f$ as
\begin{equation}
  L^2(f) = \left(\sum_{i,j,k} dx\, dy\, dz\, f_{ijk}^2\right)^{1/2},
\end{equation}
we measure the numerical error by calculating the $L^2$ norm of the error and normalizing it by the $L^2$ norm of the analytical solution:
\begin{equation}
  \text{Error} = \frac{L^2(\phi - \phi_{\text{sphere}})}{L^2(\phi_{\text{sphere}})}.
\end{equation}

\begin{figure}
  \label{fig:gravity_convergence}
\end{figure}

The results of this test are plotted in Figure \ref{fig:gravity_convergence}. We find that convergence is actually substantially better than second-order, and that the difference is largest at low resolutions. The explanation for this is that we are attempting to model a spherical object on a rectangular grid. At very low resolution, the object does not look very spherical. As the resolution is increased, the total amount of mass on the grid, as well as its location, will change as the sphere fills out. This means we are combining the true accuracy bonus from increased resolution with the artificial accuracy bonus from getting closer to solving the problem we are supposed to be solving.

We can solve one of these two sources of error by evaluating Equation \ref{eq:sphere-analytical} with a mass $M$ that corresponds to the amount of mass actually on the grid, defined as the sum of the density multiplied by the volume of each cell for all cells that have greater than ambient density. The resolution study for this case is also plotted in Figure \ref{fig:gravity_convergence}. We still obtain convergence better than second-order, indicating that we still have the geometrical problem.

The only way to fully eliminate this effect is to test a problem that does not change with resolution. The obvious companion is a cube of uniform density $\rho$, where now $R$ is half of the side length of the cube. At each resolution we use the same $R$ as for the sphere, which ensures that the cube always fills exactly the same fraction of the domain and thus has the same mass, so the only improvement comes from better sampling at higher resolution. The potential for this object has been worked out analytically by \cite{waldvogel:1976} (see also a similar result in \cite{hummer:1996}). The potential is in Equation 15 of that paper, though the last term is missing a factor of $1/2$, which destroys the symmetry. Inserting this missing factor and performing a simple coordinate transformation so that the center of the cube is at the origin, the potential is
\begin{align}
  \phi_{\text{cube}}(x,y,z) &= G\rho\sum_{i,j,k=0}^1\left(x_i y_j\, \text{tanh}^{-1}\left(\frac{z_k}{r_{ijk}}\right) + y_j z_k\, \text{tanh}^{-1}\left(\frac{x_i}{r_{ijk}}\right) + z_k x_i\, \text{tanh}^{-1}\left(\frac{y_j}{r_{ijk}}\right) \right.\notag \\
  &\left. - \frac{x_i^2}{2}\,\text{tan}^{-1}\left(\frac{y_j z_k}{x_i r_{ijk}}\right) - \frac{y_j^2}{2}\,\text{tan}^{-1}\left(\frac{z_k x_i}{y_j r_{ijk}}\right) - \frac{z_k^2}{2}\,\text{tan}^{-1}\left(\frac{x_i y_j}{z_k r_{ijk}}\right)\right)
\end{align}
where $x_0 = R + x$, $x_1 = R - x$, $y_0 = R + y$, $y_1 = R-y$, $z_0 = R+z$, $z_1 = R-z$, and $r_{ijk} = \sqrt{x_i^2 + y_j^2 + z_k^2}$. In \texttt{FORTRAN} and \texttt{C}, the inverse hyperbolic tangent is \texttt{atanh} and the inverse tangent is \texttt{atan} (\textit{not} \texttt{atan2}). This formula is valid both inside and outside the cube. The normalized $L^2$ error norm for this problem is also shown in Figure \ref{fig:gravity_convergence}, and indeed we see near-perfect second-order scaling.

The main lesson here is that in a convergence study, it is important to ensure that the physical problem does not change with resolution. Since in the case of spherical objects on rectangular grids the effect is to artificially boost convergence with resolution, in a simulation with spherical objects like stars one can envision a scenario of being fooled into believing apparently good convergence results that are simply a convolution of artificially high gravitational convergence and poor convergence in the hydrodynamics. A convergence study in this case is only fully valid if there is reason to be confident that this effect is negligible compared to other factors.

Another factor that may affect convergence testing is easily seen in Figure \ref{fig:gravity_convergence}, where at the highest resolution the error flattens out. This occurs because of a limitation in the accuracy of the multigrid solver. For problems of this size we find empirically that the multigrid solver does not converge for error tolerances smaller than about $10^{-11}$, which enforces a lower limit on the total error. This is also something to be aware of in assessing convergence for our full simulations.

\subsection{Rotation}\label{sec:rotation}

@@

A simple test to verify the Poisson solver implemented by CASTRO is
the case of gravitational free fall. In this setup, two stars, each
individually in an equilibrium state, are placed on the computational
grid, separated by an initial distance $r_0$ along the $x$ axis with
zero initial velocity. We choose stars of masses $0.6$ and $0.8\,
M_\odot$, with the lower mass star on the left side (e.g. $x < 0$) of
the domain and the higher mass star on the right, such that their
center of mass coincides with the center of the
domain. Gravitationally, the stars may be treated as point masses
until the point of contact, so the equation of motion governing the
distance $r$ between their centers of mass is that of simple free
fall:
\begin{equation}
  \ddot{r}(t) = - \frac{GM}{r},
\end{equation}
where $G$ is the gravitational constant and $M$ is the total mass of
the system. This differential equation has a closed-form solution for
the evolution time as a function of separation:
\begin{equation}
  t(r) = \sqrt{\frac{r_0^3}{2GM}} \left[ \text{arccos}\left(\sqrt{\frac{r}{r_0}}\,\right) + \sqrt{\frac{r}{r_0} \left(1 - \frac{r}{r_0}\right)}\ \right]. \label{analyticalFreeFall}
\end{equation}\MarginPar{This result is derived in the freefall directory.}
This result can be derived by recognizing that
\[
  t(r) = \int_{r_0}^r \frac{dr}{v(r)}
\]
and inserting the velocity as a function of radial separation,
\[
  v(r) = \sqrt{\frac{2GM}{r_0}\left(\frac{r_0}{r} - 1\right)}.
\]
We determine the initial separation to be consistent with the
simulation performed in Section \ref{sec:kepler}; that is, we select
an initial orbital period and use Kepler's third law to calculate the
radius of a circular orbit with that period. We select an initial
orbital period of $T = 100$ s for this simulation, so that the separation is
\[
  r_0 = 3.61 \times 10^{9}\ \text{cm}.
\]
We chose a relatively low resolution simulation to demonstrate the
capabilities of CASTRO even while using modest resources. The
computational grid is covered by a coarse grid of $48^3$ zones, with
two levels of refinement above the coarse grid. Each refined grid
carries an increase in resolution by a factor of 4 relative to the
coarser grid below it. CASTRO initially assigns $93\%$ of the domain
to be covered by the intermediate resolution grids, and $0.05\%$ of
the domain to be covered by the finest resolution grids.

@@

\subsection{Keplerian Orbit}\label{sec:kepler}
